### 2012-10-16 Docs/MESA/hopper_compile.tex
latest MESA docs

>>>>

These notes are somewhat dated because they pertain only to {\sf MESA} version 
4088. The above notes are for version 4411. In my experience, version 4088 is 
much easier to install than 4411. When using the scripts in 
{\sf AstroDev/MESA/networks/scripts}, make sure the proper list of modules to 
compile (lines containing {\tt do\_one ....}) is used. Those lists are 
version dependant.

### 2012-10-16 Docs/MESA/instructions.tex
latest MESA docs

To install {\sf MESA}, download the source code from the {\sf MESA} website, 
{\tt http://mesa.sourceforge.net}. Check the website for the latest version, 
which is 4298 with the writing of this document. The codes and algorithms 
discussed here were created when the current version of {\sf MESA} was 
version 4088. Between versions 4088 and 4298, there are minor changes that 
will cause some of the included routines to crash. To download the source code 
associated with version 4088, use the following command:
\begin{lstlisting}[language=bash,mathescape=false]
  svn co -r 4088 http://mesa.svn.sourceforge.net/svnroot/mesa/trunk mesa
\end{lstlisting}
This will download the source into a directory called {\tt mesa} located 
in the directory where this command was run. This step might take some time 
($\sim$10 min depending on the internet connection) possibly because the 
final {\tt mesa} directory will take up roughly 2.2 GB.

@@

%-------------------------------------------------------------------------
\subsubsection{Modify the Makefile (NOT using SDK)}
The main makefile that contains compiler information including flags is 
located in the {\tt mesa/utils} directory and is called 
{\tt makefile\_header\_non\_mesasdk}. Open it and change the following:
\begin{enumerate}
\item Compilers: First change the compilers labeled {\tt CC} (C compiler) 
and {\tt FC} (Fortran compiler) to the compilers of your choice ({\sf MESA} 
defaults are {\tt gcc} and {\tt ifort}).

@@

\item {\tt PGPLOT}: Set the variable {\tt USE\_PGSTAR} to {\tt NO} and set 
the variable {\tt LOAD\_PGPLOT} to be empty on the line corresponding to 
the compiler that was chosen. The default for {\tt gcc} and {\tt ifort} is 
{\tt YES}, but for any other compiler the default is {\tt NO}. {\tt PGPLOT} 
controls the pop up images that are used in {\tt mesa/star} (1D stellar 
evolution code).

\item {\tt USE\_SE}: Set the variable {\tt USE\_SE} to {\tt NO} and set both 
{\tt LOAD\_SE} and {\tt INCLUDE\_SE} to be empty. {\tt SE} is a library of 
formats used for writing/reading data in stellar evolution codes.

\item Compiler Flags: Installing {\sf MESA} without OpenMP will result in a 
failed install. {\sf MESA} is capable of being compiled in serial, but it 
must happen after the initial install. If not using {\tt gcc} or {\tt ifort} 
it will be necessary to enter all of the relevant compiler flags.
\end{enumerate}

@@

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Handy compilation script
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Compilation Script}
 \label{sec:compile_scripts}

Useful script written in {\tt BASH} to help with compiling {\sf MESA} modules. 
The script cleans the module directory, compiles the source code and exports 
the libraries and {\tt .mod} files to the appropriate directories for every 
module. This script can be found in {\tt AstroDev/networks/MESA/scripts}. 
There is also a script to compile just one module. The compile-all-modules 
script should be run from the top {\sf MESA} directory while the 
compile-one-module script can be run from anywhere because it takes the 
relative path to the module.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Script
%%%%%%%%
\subsection{Compile All Modules}
\input compile_all_mods

>>>>

To install {\sf MESA}, download the source code from the {\sf MESA} website, 
{\tt http://mesa.sourceforge.net}. Check the website for the latest version, 
which is 4442 with the writing of this document. The codes and algorithms 
discussed here were created when the current version of {\sf MESA} was 
version 4411. Between versions 4411 and 4442, there are minor changes that 
will cause some of the included routines to crash. To download the source code 
associated with version 4411, use the following command:
\begin{lstlisting}[language=bash,mathescape=false]
  svn co -r 4411 http://mesa.svn.sourceforge.net/svnroot/mesa/trunk mesa
\end{lstlisting}
This will download the source into a directory called {\tt mesa} located 
in the directory where this command was run. This step might take some time 
($\sim$10 min depending on the internet connection) possibly because the 
final {\tt mesa} directory will take up roughly 2.2 GB.

@@

%-------------------------------------------------------------------------
\subsubsection{Modify the Makefile (NOT using SDK)}
The main makefile that contains compiler information including flags is 
located in the {\tt mesa/utils} directory and is called 
{\tt makefile\_header\_non\_mesasdk}.

All of the following changes apply to 
variables that already exist. There is no need to explicitly add new code 
unless you intend to compile {\sf MESA} with something other than 
{\tt gfortran} or {\tt ifort}. If you want to compile using something 
different, then the only code that needs to be explicitly added are the 
definitions of the compiler flags.

Open the {\tt makefile\_header\_non\_mesasdk} and change the following:
\begin{enumerate}
\item Compilers: First change the compilers labeled {\tt CC} (C compiler) 
and {\tt FC} (Fortran compiler) to the compilers of your choice ({\sf MESA} 
defaults are {\tt gcc} and {\tt ifort}).

@@

\item {\tt PGPLOT}: Set the variable {\tt USE\_PGSTAR} to {\tt NO} and set 
the variable {\tt LOAD\_PGPLOT} to be empty in the if-construct corresponding 
to the compiler that was chosen. The default for {\tt gcc} and {\tt ifort} is 
{\tt YES}, but for any other compiler the default is {\tt NO}. {\tt PGPLOT} 
controls the pop up images that are used in {\tt mesa/star} (1D stellar 
evolution code).

\item {\tt USE\_SE}: Set the variable {\tt USE\_SE} to {\tt NO} and set both 
{\tt LOAD\_SE} and {\tt INCLUDE\_SE} to be empty. These also need to be set 
in the correct if-construct according to which compiler was chosen. 
{\tt SE} is a library of 
formats used for writing/reading data in stellar evolution codes.

\item Compiler Flags: Find the if-construct corresponding to the chosen 
compiler. This is where compiler options are defined. It is not necessary to 
edit any of these if you are using {\tt gfortran} or {\tt ifort}. These are 
the variables to edit if you wish to add more (or less) debugging flags 
for example. If using a different compiler, the variables in the last 
{\tt else} statement (which are all empty except for 
{\tt FCbasic = UNKNOWN COMPILER}) are what must be edited to hold the compiler 
flag definitions.

Installing {\sf MESA} without OpenMP will result in a 
failed install. {\sf MESA} is capable of being compiled in serial, but it 
must happen after the initial install.

\item {\tt LOAD\_MATRIX}: {\tt LOAD\_MATRIX} is first defined near the end of 
the file around line $\sim$275. The purpose of this variable is to load all 
the libraries associated with linear algebra including {\tt LAPACK}, {\tt BLAS} 
and a few others. The lines we wish to change look like this:
\begin{lstlisting}[language=make,mathescape=false]
  #LOAD_MATRIX = -lmtx -lconst -lmesaklu $(LOAD_LAPACK) $(LOAD_BLAS)
  # for ifort on mac, we need the following def for LOAD_MATRIX 
  # because of loader bugs (as of July 2012)
  LOAD_MATRIX = -lmtx $(MESA_DIR)/const/make/const_def.o \
      $(MESA_DIR)/const/make/const_lib.o -lmesaklu $(LOAD_LAPACK) $(LOAD_BLAS)
  LOAD_MESA_NUMERICS = -linterp_2d -linterp_1d -lnum -lutils -lalert \
      $(LOAD_MATRIX)
\end{lstlisting}
We must uncomment the first {\tt LOAD\_MATRIX} and comment out the second one:
\begin{lstlisting}[language=make,mathescape=false]
  LOAD_MATRIX = -lmtx -lconst -lmesaklu $(LOAD_LAPACK) $(LOAD_BLAS)
  # for ifort on mac, we need the following def for LOAD_MATRIX 
  # because of loader bugs (as of July 2012)
  #LOAD_MATRIX = -lmtx $(MESA_DIR)/const/make/const_def.o \
  #    $(MESA_DIR)/const/make/const_lib.o -lmesaklu $(LOAD_LAPACK) $(LOAD_BLAS)
  LOAD_MESA_NUMERICS = -linterp_2d -linterp_1d -lnum -lutils -lalert \
       $(LOAD_MATRIX)
\end{lstlisting}

\item {\tt install} script: In the {\tt MESA} directory, open the {\tt install} 
script and comment out two lines. We want to change:
\begin{lstlisting}[language=bash,mathescape=false]
  do_one atm

  do_one sample

  do_one star

  do_one adipls
\end{lstlisting}
to this:
\begin{lstlisting}[language=bash,mathescape=false]
  do_one atm

  #do_one sample

  #do_one star

  #do_one adipls
\end{lstlisting}
There are routines in the {\tt star} package and {\tt adipls} that rely on 
the {\tt PGSTAR} modules. We do not wish to use these modules and the easiest 
way around this is to not install both {\tt star} and {\tt adipls} because 
neither one is needed to include {\sf MESA} microphysics modules in 
{\sf MAESTRO}. {\tt star} is the 1D stellar evolution module and {\tt adipls} 
module is used for asteroseismology. The {\tt sample} module is simply a 
sample directory that is not needed for using the main microphysics modules.

\item Edit 1 source file: An undefined reference to 
{\tt omp\_get\_max\_threads} occurs in the {\tt const\_def.f} file located in 
{\tt const/public/const\_def.f}.

In the {\tt const\_def.f} file, around line $\sim$164, in the 
{\tt do\_const\_init} subroutine, there is a line that reads:
\begin{lstlisting}[language=fortran,mathescape=false]
  omp_num_threads = omp_get_max_threads()
\end{lstlisting}
Our first change is to declare a temporary character variable in the 
({\tt do\_const\_init} subroutine):
\begin{lstlisting}[language=fortran,mathescape=false]
  character(len=16) :: temp_max_threads
\end{lstlisting}
Then we comment out the above line ({\tt omp\_num\_threads = ...}) and add the 
following (all in the {\tt do\_const\_init} subroutine):
\begin{lstlisting}[language=fortran,mathescape=false]
  !omp_num_threads = omp_get_max_threads()
  call get_environment_variable("OMP_NUM_THREADS", temp_max_threads)
  read(temp_max_threads, *) omp_num_threads
\end{lstlisting}

\end{enumerate}

@@

The routines that carry out these processes (setting up EOS, setting up 
isotope map, shutdown routines, etc.) have already been written and live in 
{\sf AstroDev/networks/MESA} and {\sf AstroDev/EOS/MESA}. This section 
outlines the necessary steps taken by those routines in order to set up the 
EOS and network.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Handy compilation script
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Compilation Scripts}
 \label{sec:compile_scripts}

Some useful scripts have been written in {\tt BASH} to help with compiling 
{\sf MESA} modules. They live in the {\sf AstroDev/networks/MESA/scripts} 
directory. Currently, there are two scripts; one that compiles all modules 
(which unfortunately is version dependant), and one that compiles a single 
module.

The compile-one-module script can be run from any directory because the 
calling sequence involves the relative path to the {\sf MESA} module:
\begin{lstlisting}[language=bash,mathescape=false]
  compile-one-mod.sh ../../../mesa/net
\end{lstlisting}
The above command will compile the {\tt net} module without needing to be in 
the {\sf MESA} directory.

The compile-all-modules script will clean the module directory, compile the 
source code and export the libraries and {\tt .mod} files to the appropriate 
directories for every module. To run this script, you must be in the {\sf MESA} 
directory. The calling sequence would be something like:
\begin{lstlisting}[language=bash,mathescape=false]
  cd $MESA_DIR
  compile-all-mods.sh
\end{lstlisting}
